Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-13 paper-ieee-trans.tex

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
supporting an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One such example is the Worldwide 
Large Hadron Collider (LHC) Computing Grid infrastructure, motivated 
by the particle physics scientific community, build to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the 5th, 6th and 7th Framework Programmes,
and recently by the Horizon 2020 programme \cite{h2020} -- are under way to
develop software solutions that provide a seamless and transparent access to distributed
computing and data resources by the researchers.

@@

Soon it become apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle would have to be devoted to the quality of the software being
delivered. The evolution of software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of teams
and deployment of pilot testbeds.

Analysing the European Commission (EC)-funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. One of many reasons, is related to the
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization). All of which
complemented by the emergence of automation and event-response tools.

In this paper, we outline a set of EC-funded projects
ranging from 2001 to now, in which the authors were involved.
For each project, the challenges, the methodologies and
the achievements towards software reliability are described.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way of the subsequent projects,
following a logic evolution in which the role of the quality and reliability of the software have
consistently increased. The experience gained during the 
course of each project support this fact as shown in Table \ref{tab:feat}.
The most recent projects are more committed and aware to software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

The {\sl gLite} middleware \cite{glite} was the
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with the {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, fixing more than 10 thousand bug, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National 
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops}, 
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the 
management of the software provisioning process \cite{mario}, to validate and distribute the software to
the production infrastructure. As a result, this process led to a production--ready 
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). The {\sl UMD}
was a stack of about 250 software components from several technology providers such 
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The 
QC defines a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper thus, it
emerged from the lessons learned in the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of
large collaborative projects with an extensive list of partners.

@@

The software quality policies were initially defined in the first deliverable document of the
project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
from the user communities and feedback from the development teams. These policies cover the 1) identification
and description of the \emph{SQA requirements} that the produced software need 
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's 
behaviour throughout the development, release and post--release stages.

>>>>

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
supporting an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, motivated
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the 5th, 6th and 7th Framework Programmes,
and recently by the Horizon 2020 programme \cite{h2020} -- are under way to
develop software solutions that provide a seamless and transparent access to distributed
computing and data resources by the researchers.

@@

Soon it become apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle would have to be devoted to the quality of the software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of teams
and deployment of pilot testbeds.

Analysing the European Commission (EC)-funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. The
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization) definitely motivated this trend. All of which
complemented by the emergence of automation and event-response tools.

In this paper, we outline a set of EC-funded projects
ranging from 2001 to nowadays, in which the authors were involved.
For each project, the challenges, the methodologies and
the achievements towards software reliability are described.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way of the subsequent projects,
following a logic evolution in which the role of the quality and reliability of the software have
consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed and aware to software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

The {\sl gLite} middleware \cite{glite} was the
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with the {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, which issued more than 10 thousand bug, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops},
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
management of the software provisioning process \cite{mario}, to validate and distribute the software to
the production infrastructure. As a result, this process led to a production--ready
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
was a stack of about 250 software components from several technology providers such
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
QC defines a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

@@

The software quality policies were initially defined in the first deliverable document of the
project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies cover the 1) identification
and description of the \emph{SQA requirements} that the produced software need
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's
behaviour throughout the development, release and post--release stages.
